Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-05 paper/paper.tex

% text macros
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\panstarrs}{\project{Pan\acronym{STARRS1}}}
\newcommand{\ps1}{\project{\acronym{PS1}}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

Here we build and use just such a data-driven model to infer more precise parallaxes for each star.
In particular, we use all of the \gaia\ \tgas\ data (\citealt{tgas};
these data were generated by the methodology of \citealt{michalik15}), that match to the
\tmass\ photometry (\citealt{skrutskie06}) and lie in the footprint of \panstarrs\ (\ps1)
to make use of the \cite{green15} dust map. We use this data to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We build the \cmd\ using an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its disadvantage, for the present purposes, is that it requires
or implies a strictly Gaussian noise model.
As we show below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical Bayes, or maximum-marginalized likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \ps1-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

We use stars crossmatched in \tgas\ and \tmass\.
The match was done using a nearest-neighbor algorithm with a search radius of $4"$.
\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}
We also required that the stars lie within the
observing footprint of \ps1\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia, and the $1\sigma$ and
$2\sigma$ contours of the distribution.
The right panel shows a subset of the data with it's associated error bars.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

In detail, we take the $5\%$
quantile of each distance posterior (the closest part of the
distance posterior) and query the 3D dust map at that distance.
We take the $5\%$ quantile---as apposed to the median---to do a minimal dust
correction because over-correcting for dust can push the peak of the likelihood
function into an unphysical, different space of the \cmd\ and thus bias the
parallax inference.
For each distance, we sample the median of the
probabilistic dust model. We apply each dust correction to our
\tmass\ photometry using Equation~\ref{eq:data}, where $Q_{\lambda}
= [0.709, 0.302]$ for bands $[J, K]$ respectively (see Table 6 in
\citealt{schlafly11}).
We then regenerate our prior using the dust-corrected photometry.
The dust values we obtain in our final
iteration, and which we apply to our photometry, are visualized in
galactic coordinates in \figurename~\ref{fig:dust}.
After applying the dust correction, we do not update the uncertainties, which is
technically wrong and under-represents the true uncertainties; as mentioned
previously, this leads to a less severe deconvolution.
In other words, the features of our inferred
  \cmd\ are conservatively broad (broader than if we over-estimated
  the errors which would have lead to excessive deconvolution and
  unrealistically narrow features). We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept. We iterate this process 10 times, however
the dust values seem to converge after about 2-3 iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance quantile in Galactic coordinates. This shows both the footprint of our analysis as well as the adopted extinction values. Each point represents a single star in our cross matched catalog. The large missing areas are due to the \ps1\ footprint ($>-30~{\rm deg}$ in declination)}
\label{fig:dust}
\end{figure}

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Jo Bovy (Toronto),
  Andy Casey (Monash),
  Stephen Feeney (Flatiron),
  Dustin Lang (Toronto),
  David Spergel (Flatiron),
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

This work has made use of data from the European Space Agency (\acronym{ESA})
mission \gaia\footnote{\url{http://www.cosmos.esa.int/gaia}}, processed by the Gaia
Data Processing and Analysis Consortium\footnote{\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}} (\acronym{DPAC}). Funding for the
\acronym{DPAC} has been provided by national institutions, in particular the
institutions participating in the Gaia Multilateral Agreement.

@@

%% \appendix
%% \section{Appendix material}

@@

\clearpage

@@

%%
%% End of file `sample.tex'.

>>>>

% text macros
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\panstarrs}{\project{Pan\acronym{STARRS}}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

Here we build and use just such a data-driven model to infer more precise parallaxes for each star.
In particular, we use all of the \gaia\ \tgas\ data (\citealt{tgas};
these data were generated by the methodology of \citealt{michalik15}), that match to the
\tmass\ photometry (\citealt{skrutskie06}) and lie in the footprint of
the \panstarrs\ Dust Map (\citealt{green15}), to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We build the \cmd\ using an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its disadvantage, for the present purposes, is that it requires
or implies a strictly Gaussian noise model.
As we show below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical Bayes, or maximum-marginalized likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \panstarrs-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

We use stars crossmatched in \tgas\ and \tmass\, that also lie within the
observing footprint of \panstarrs\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia, and the $1\sigma$ and
$2\sigma$ contours of the distribution.
The right panel shows a subset of the data with it's associated error bars.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

In detail, we take the $5\%$
quantile of each distance posterior (the closest part of the
distance posterior) and query the 3D dust map at that distance.
We take the $5\%$ quantile---as apposed to the median---to do a minimal dust
correction because over-correcting for dust can push the peak of the likelihood
function into an unphysical, different space of the \cmd\ and thus bias the
parallax inference.
For each distance, we sample the median of the
probabilistic dust model. We apply each dust correction to our
\tmass\ photometry using Equation~\ref{eq:data}, where $Q_{\lambda}
= [0.709, 0.302]$ for bands $[J, K]$ respectively (see Table 6 in
\citealt{schlafly11}).
We then regenerate our prior using the dust-corrected photometry.
The dust values we obtain in our final
iteration, and which we apply to our photometry, are visualized in
galactic coordinates in \figurename~\ref{fig:dust}.
After applying the dust correction, we do not update the uncertainties, which is
technically wrong and under-represents the true uncertainties; as mentioned
previously, this leads to a less severe deconvolution.
In other words, the features of our inferred
  \cmd\ are conservatively broad (broader than if we over-estimated
  the errors which would have lead to excessive deconvolution and
  unrealistically narrow features). We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept. We iterate this process 10 times, however
the dust values seem to converge after about 2-3 iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance quantile in Galactic coordinates. This shows both the footprint of our analysis as well as the adopted extinction values. Each point represents a single star in our cross matched catalog. The large missing areas are due to the \panstarrs\ footprint ($>-30~{\rm deg}$ in declination)}
\label{fig:dust}
\end{figure}

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Jo Bovy (Toronto),
  Andy Casey (Monash),
  Stephen Feeney (Flatiron),
  Dustin Lang (Toronto),
  David Spergel (Flatiron),
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} \gaia\ Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

This work has made use of data from the European Space Agency (\acronym{ESA})
mission \gaia\footnote{\url{http://www.cosmos.esa.int/gaia}}, processed by the \gaia\ Data
Processing and Analysis Consortium\footnote{\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}} (\acronym{DPAC}). Funding for the
\acronym{DPAC} has been provided by national institutions, in particular the
institutions participating in the \gaia\ Multilateral Agreement.

@@

\software{
The code used in this project is available from
\url{https://github.com/andersdot/photoParallax}.
This research utilized the following open-source \textsl{Python} packages:
    \textsl{Astropy} (\citealt{Astropy-Collaboration:2013}),
    \textsl{matplotlib} (\citealt{Hunter:2007}), and
    \textsl{numpy} (\citealt{Van-der-Walt:2011}).
This work additionally used the \gaia\ science archive
(\url{https://gea.esac.esa.int/archive/}).
}

@@

@@
